---
title: "lab_03"
author: "Cecilia Clemens"
date: "2025-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)
# Load janitor
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=369008) |>
  clean_names() |> 
  mutate(source = str_to_upper(vendor_name))
# Load in the Maryland grants and loan data
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv", guess_max=19482)|> 
  clean_names() |>
  mutate(description = str_to_upper(description))

```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1:I found 81 records for payments for payments to the Prince George's County School system.

```{r}
pg_schools <- payments |> filter ( str_detect(vendor_name, "PRINCE GEORGE'S COUNTY PUBLIC SCHOOL") | str_detect (vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOL")| str_detect(vendor_name, "PRINCE GEORGES COUNTY SCHOOL") | str_detect(vendor_name, "PRINCE GEORGE'S COUNTY SCHOOL") | str_detect(vendor_name, "PGCPS")| str_detect(vendor_name, "PG COUNTY PUBLIC SCHOOL") | str_detect(vendor_name, "PRINCE GEORGES PUBLIC SCHOOL") | vendor_name == "BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD")
nrow(pg_schools)
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: The year 2024 had the largest total amount with $1,782,519,925 given to the school system. 

```{r}
pg_schools |>
  group_by(fiscal_year) |>
  summarise(
    count_totals = sum(amount)
  )
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. Unfortunately, many of these vendors don't have a standardized way of spelling their names and are also prone to spelling errors which means I will have to look for ways to assign the data to only one name. The names A & I inc, A&I Inc, and A&I Inc. are all the same company. Another example is A Morton Thomas & Associates Incorporated which abbreviated words differently each time they put in their names. The company Aerosol Monitoring & Analysis Incorporated misspelled Aerosol as Aersol. I would use the case_when function to assign the names to a standardized version of each vendor name. 

```{r}
UMD_payments <- payments |> filter(str_detect(agency_name, "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND") )

UMD_payments |>
 group_by(vendor_name) |>
  summarise(
    count_totals = sum(amount)
  )


```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: Since 2009 stem cell research funding in Maryland has been steadily decreasing but a new surge in funding in 2023 has revitalized research. 
```{r}
stem_cell_research <- md_grants_loans |> 
  filter( str_detect(description, "STEM CELL"))
stem_cell_research |>
  group_by(fiscal_year) |>
  summarise(
    yearly_total = sum(amount)
  )

```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: There are 18 rows of grantees in total that the code shows but looking at the data I see that the University of Maryland is listed three different ways. So there are actually 16 grantees in this zip code. TO clean up the data I would just have to compile the amounts listed under the three different UMD names and assign it to one grantee name. The grantee that stands out to me is the UMD Office of the Comptroller. I assumed it would be the office in charge of the money going in and out of the University. I think it's always important to see how the people in charge of money are using their money. A more than $400,00 loan is a significant amount to be awarded to an office that is not involved with research. I then went and looked up the office and the only website they have is one for something called the "Kuali Financial System." There may be a simple explanation for this but out of all the grantees that is the one that made the least sense to me. 

```{r}
UMD_grants_loans <- md_grants_loans |>
  filter(str_detect(zip_code, "20742"))|>
  group_by(grantee) |>
  summarise(
    total_amount = sum (amount)
  )

UMD_grants_loans

  
```
